---
title: "Beech Leaf Disease and Dendro-cores"
author: "Nathan Byer, Connie Hausman, Katie Flinn"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    fig_width: 6
    fig_height: 4
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,out.width = "100%")

options(width=1200)


detachAllPackages <- function() {
  basic.packages.blank <- c(    
    "stats",    
    "graphics",    
    "grDevices",    
    "utils",   
    "datasets",  
    "methods",    
    "base"    
  )    
  basic.packages <- paste("package:", basic.packages.blank, sep = "")   
  package.list <- search()[ifelse(unlist(gregexpr("package:", search())) == 1, TRUE, FALSE)]   
  package.list <- setdiff(package.list, basic.packages)   
  if (length(package.list) > 0) {   
    for (package in package.list) {   
      detach(package, character.only = TRUE)   
    }   
  }    
}

detachAllPackages()
library(ggplot2)
library(dotwhisker)
library(dplyr)
library(tidyverse)
library(vegan)
library(caret)
library(lubridate)
library(spOccupancy)
library(vegan)
library(raster)
library(dismo)
library(terra)
library(sp)
library(rgdal)
library(gstat)
library(broom)
library(shiny)
library(sp)
library(ggplot2)
library(RColorBrewer)
library(rgdal)
library(broom)
library(AICcmodavg)
library(indicspecies)
library(lme4)
library(MASS)
library(cooccur)
library(igraph)
library(visNetwork)
library(lubridate)
library(performance)
library(glmmTMB)
```

# Reading in the tree ring data

```{r}
beech<-read.csv("beech_master.csv",header=T)

hemlock<-read.csv("hemlock_master.csv",header=T)

maple<-read.csv("maple_master.csv",header=T)


```

# Reading in BLD data

```{r}
bld_data<-read.csv("2022_BWU_Flinn_tree_rings_BLD_datarequest.csv",header=T)
```

# Merging all three, and adding a key for tree species

```{r}
beech_transpose<-beech %>%
  pivot_longer(cols = ABW.101A:ABW.301A,names_to = "Tree") %>%
  add_column(Species="F. grandifolia") 

hemlock_transpose<-hemlock %>%
  pivot_longer(cols = ABW.201A:ABW.300B,names_to = "Tree") %>%
  add_column(Species="T. canadensis")

maple_transpose<-maple %>%
  pivot_longer(cols = ABW.100A:ABW.9B.1,names_to = "Tree") %>%
  add_column(Species="A. spp.")

alltrees<-rbind(beech_transpose,hemlock_transpose,maple_transpose)

p<-ggplot(aes(x=Year,y=value,col=Species),data=alltrees)+geom_point()+geom_smooth()+ylim(c(0,20))
p

p2<-ggplot(aes(x=Year,y=value,col=Species),data=alltrees)+stat_summary(geom = "line")+ylim(c(0,20))
p2

p3<-ggplot(aes(x=Year,y=value,col=Species),data=alltrees)+stat_summary(geom = "line")+ylim(c(0,20))+xlim(c(2000,2019))
p3

```

## create periods for pre- during - post

```{r}
alltrees$BLDperiod<-as.numeric(alltrees$Year)

alltrees$BLDperiod[alltrees$BLDperiod < 2012]<- "before"

alltrees$BLDperiod[alltrees$BLDperiod != "before"]<- "after"


```

## models of species x period

```{r}


alltrees$BLDperiod<-factor(alltrees$BLDperiod,levels=c("before","after"))
alltrees$Species<-factor(alltrees$Species,levels=c("F. grandifolia","A. spp.","T. canadensis"))

growthmod<-glmmTMB(value ~ BLDperiod*Species+
                           (1|Tree), 
                       data = alltrees)

summary(growthmod)

```

So this suggests that beech trees typically grow more slowly than maples, and faster than hemlocks. However, for the post-BLD period, beech trees have actually *increased* in growth rates overall, and now appear to be dramatically faster growing relative to maples and hemlocks (based on the significant, negative interaction terms). 

1. pull in BLD prevalence
2. explore beech-specific associations between growth rates and BLD 
3. compare growth rates for BLD-infected beeches to non-beech trees


#  

```{r}

treename<-gsub("ABW.", "", beech_transpose$Tree)
treename<-gsub("A", "", treename)
treename<-gsub("B", "", treename)

beech_transpose$tree_id<-as.numeric(treename)

beech_transpose$treeyear<-paste0(beech_transpose$tree_id,"_",beech_transpose$Year)

bld_data$Year<-year(as.Date(bld_data$date,format = "%m/%d/%Y"))
bld_data$treeyear<-paste0(bld_data$tree_id,"_",bld_data$Year)

bld_data_symptomdata<-bld_data[,c(2,10:21,23:25)]

bld_data_symptomdata_naomit<-na.omit(bld_data_symptomdata)

cols<-seq(2,16,1)

bld_data_symptomdata[,cols] = apply(bld_data_symptomdata[,cols], 2, function(x) as.numeric(as.factor(x)))

bld_data_symptomdata_2<-column_to_rownames(bld_data_symptomdata,var = "tree_id")

bldrda<-rda(bld_data_symptomdata_2)

scores(bldrda,display = "species")


# PC1 captures a sick (negative) to healthy (positive) gradient
# will just use PC1 scores for each tree

bld_data$PC1<-scores(bldrda,display="sites")[,1]

bld_data$relhealth<-bld_data$PC1

bld_data$relhealth[bld_data$relhealth < 0]<- "unhealthy"

bld_data$relhealth[bld_data$relhealth != "unhealthy"]<- "healthy"

beech_transpose_2 <- beech_transpose %>%
  group_by(tree_id,treeyear,Year) %>%
  summarise(growthincrement=value) %>%
  left_join(bld_data,by="treeyear") 

beech_transpose_3<-beech_transpose_2[beech_transpose_2$Year.x %in% c(seq(2000,2021,1)),]


beech_transpose_3$BLDperiod<-as.numeric(beech_transpose_3$Year.x)

beech_transpose_3$BLDperiod[beech_transpose_3$BLDperiod < 2012]<- "before"

beech_transpose_3$BLDperiod[beech_transpose_3$BLDperiod != "before"]<- "after"


bld_data_health<-bld_data %>%
  dplyr::select(tree_id,relhealth,PC1)
bld_data_health$tree_id.x<-bld_data_health$tree_id

beech_transpose_4<- beech_transpose_3 %>%
  group_by(tree_id.x,Year.x,BLDperiod) %>%
  summarise(
            averagegrowth=mean(growthincrement)) %>%
  drop_na() %>%
  left_join(bld_data_health,by="tree_id.x")

```


```{r}
beech_transpose_4
growthhealthmod<-glmmTMB(averagegrowth ~ PC1*BLDperiod+
                           (1|tree_id.x), 
                       data = beech_transpose_4)

summary(growthhealthmod)
```

```{r}
beech_transpose_5<- beech_transpose_3 %>%
  group_by(tree_id.x) %>%
  summarise(
            averagegrowth=mean(growthincrement)) %>%
  drop_na() %>%
  left_join(bld_data_health,by="tree_id.x")

growthhealth_end<-glm(averagegrowth ~ PC1,
                       data = beech_transpose_5)

p<-ggplot(aes(x=PC1,y=averagegrowth),data=beech_transpose_5)+geom_point()+geom_smooth(method="lm")+xlab("PC-based BLD health score (negative = more symptoms)") + ylab("average growth rate (cm, 2000-2021)")
p
```

